### 2015-04-20 Paper1rev2/paper1rev1.tex
updated for rev2

\usepackage[section]{placeins}

@@

\subsection*{Intent Extraction: Deducing the Desired Trajectory}
We first demonstrate the \textit{intent extraction} approach in human motor control, but we later show that the process is applicable to any controlled process with an invertible highest order plant term. The process begins by presuming a model of the controller. Here, we choose the well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} where the feedforward aspect of the controller perfectly predicts the plant and linearizes the system through cancellation. Additionally, linear feedback rejects position and velocity error. This model was chosen to help illuminate the approach, but later we present how to generalize this approach to more complex models without loss of generality. The equation governing the passive planar dynamics (plant) of musculoskeletal structure is of the form,
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=0
\end{equation}
where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocity, $\ddot{q}$ is joint angular acceleration, $G$ contains both Coriolis and centripetal effects, and $E$ is any externally-applied torque. The motion behavior changes with the addition of feedforward and/or feedback controllers,
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=\underbrace{\overbrace{\hat{M}(q_d)\ddot{q}_d}^{\text{Inertia}}+\overbrace{\hat{G}(q_d,\dot{q}_d)}^{\text{Coriolis, Centripal}}+\hat{E}}_{\text{Feedforward Controller}}+\underbrace{K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})}_{\text{Impendance, Feedback Controller}}
\end{equation}
where terms with hats over them ($\hat{M}$, $\hat{G}$, or $\hat{E}$) indicate that they represent the nervous system's best estimate of the forces and dynamics it will encounter, which is also known as an internal model \cite{shadmehr1994adaptive}. This portion of the system serves as an inverse-dynamics feedforward controller that cancels out the dynamics of the arm in the torque balance. If the nervous system has sufficient experience and is expecting $E$, it is included as part of the internal model, $\hat{E}$; otherwise $\hat{E}$ is set to zero. $K_p$ and $K_d$ are the lumped impedance and feedback terms that employ a moving state equilibrium to accomplish the desired trajectory, $q_d$. This $q_d$ has a dual meaning in that it signifies both the unknown desired trajectory that we seek to discover and also the moving equilibrium trajectory of the arm.

@@

To explore the conditions under which this procedure for recovering the intent trajectory can be generalized, we take a series of steps to leverage a relationship between the equilibria of a nonlinear system and the equilibria of linearizations of that system. First, we distinguish between joint space and generalized space by introducing a generalized state variable, $x$. Second, we separate the plant (arm) into its process (dynamics), $P$, and actuator (muscles), $A$, components. Third, we model these components in a very general sense as time-varying linear systems,
\begin{equation} \label{eq:generalPlant}
\underbrace{\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+\overbrace{\sum_{m=1}^M A_m(x-x_e)^{(m)}}^\text{Actuator}}_\text{Plant}+E=0
\end{equation}
where $x_e$ is the equilibrium trajectory of the actuator. Linearization is informative here because stable equilibrium points of linearizations of nonlinear systems are also equilibrium points of the nonlinear system\cite[p.~156]{khalil2002nonlinear} as long as the Jacobian is bounded and has a bounded time derivative (Lipschitz continuous)\cite[p.~161]{khalil2002nonlinear}. Fourth, we describe a  control law for the actuator using the same expansions,
\begin{equation}
\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m(x_d-x_e)^{(m)}}^\text{Actuator}=0
\end{equation}
where an internal model, $\hat{P}$ and $\hat{E}$, predicts system dynamics and external disturbances in order to determine the actuator equilibrium $x_e$ such that $x$ will track a desired path $x_d$. The feedforward component must use the actuator and hence it shares the actuator's equilibrium with the plant. This allows us to finally relate the physical system and its control law by solving for $x_e$ and substituting into \eqref{eq:generalPlant},
\begin{equation}
\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+E=\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m (x_d-x)^{(m)}}^\text{Feedback (Actuator)}
\end{equation}
wherein $x_e$ vanishes recovering our familiar model. While a proper choice of $x_e$ is perhaps essential for control, our approach does not require a model relating this equilibrium and intended trajectory. Note also that the actuator's equilibrium $x_e$ is not the process's equilibrium unless all derivatives of $x_d$ are zero. This is important as the actuator state ($\lambda$) has been claimed to be the intended movement\cite{feldman1995origin}. This upholds Gomi and Kawato's finding that the arm muscles' equilibrium does not represent reaching intent\cite{gomi1997human}. If the highest order coefficient of the dynamic model of the process $\hat{P}_N$ can be inverted, the impedances can be modeled, and the linearized system is stable, it is possible to solve for $x_d^{(N)}$ and integrate for $x_d$, revealing the intended trajectory.

While we treat both the process and actuator as dynamic models in our derivation, it is important to note that both the derivation and the technique are agnostic to the process model so long as its input is a stable equilbrium. Instead of a dynamic model, it could be a lookup table. For instance, a switch is well-modeled as a lookup table or threshold without any consideration of its underlying mechanism. In this case, our technique could be used to determine a person's intention to flip the switch and how it changes in the face of disturbance. As long as some bidirectional relationship exists between state and outcome (even if determined empirically), intended outcome can be determined. This allows determination of intent in many situations of interest, even where the process is otherwise irreducibly complex.

@@

\subsection*{Indices of variance-based sensitivity}
To ensure a properly spaced and efficient evaluation of sensivity, we employed Sobol-distributed matrices, generated using MATLAB's \textit{sobolset()} function and converted to parameter distributions (see Table~\ref{tab:parameters}) using inverse cumulative probability density. Nominal parameter values were mostly taken from Dempster \cite{dempster1955space} and Burdet et al. \cite{burdet2006stability} with typical values for subject-specific parameters (height and weight) and direct measurements for device-specific parameters (force sensor noise and drift). Standard deviations reflect expected variation in repeated measurement (ie. height and weight) or measured variance (ie. trial-to-trial variation in sensor noise or uncertainty in mass ratios reported by Dempster \cite{dempster1955space}). Distributions were clamped to the range of $\pm3$ standard deviations in order to limit parameters to a realistic range. Combinations and calculations were made as prescribed by Saltelli et al. \cite{saltelli2010variance} to arrive at direct and total sensitivity indices.

@@

Nevertheless, there are some limitations to these methods. Accurate interface force measurement (within a few tenths of a newton) is needed. Such instrumentation is available, but must be used with great care in order to preserve the accuracy of the estimate. Our sensitivity analysis on synthetic data revealed that slight error can lead to a trajectory that accumulates a drift due to bias caused by the force term. More importantly, while modern force sensors can be highly accurate with very high signal to noise ratios, they tend to drift over time, leading to error that can grow if the device is not periodically tared. Our empirical approach used several methods to mitigate these effects, and hence the errors became small. It remains to be seen if there are methods to further reduce the mis-estimation due to force error or eliminate the need for force estimation altogether.

A key assumption in this study is that the joint impedance (stiffness) is linearly related to torque. While other approaches tend to construct paradigms that indicate or presume that the intention is static in order to derive impedance -- the so called family of system identification applications, this study makes a constrained assumption of stiffness in order to determine intent. In a sense, this approach widens the scope to include movement. Sensitivity analyses showed that approximation of stiffness is the most critical factor to accuracy in estimating intent. While serious trial-to-trial stiffness mis-estimation is unlikely, co-contraction in response to disturbance might plausibly double stiffness mid-reach. Even as the model is insensitive to normal variations, a human can halve or double their stiffness \cite{franklin2007endpoint} significantly reducing determination accuracy. This stiffening may even explain the reported deviation of intent from the baseline trajectories. These limitations are readily overcome by performing this sensitivity analysis on each force environment of interest and taking measures to detect co-contraction and hence stiffness alterations. Additionally, real-time determination of intent may allow real-time determination of stiffness.

@@

% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
Research supported by US NIH NINDS1 R01 NS053606.  We thank Felix Huang, Jagan Jimmy, Christine Massie, Tejas Madhavan, and Mohi Ahmed for their review of preliminary manuscript. We also thank the community  of the Robotics Lab at the Rehabilitation Institute of Chicago for critical commentary during the genesis of this paper and the work leading up to it.

%\section*{References}
\bibliography{paper1.bib}
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% OR
%
% Compile your BiBTeX database using our plos2009.bst
% style file and paste the contents of your .bbl file
% here.
%
\pagebreak

\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\caption{
{\bf Simulated data illustrating tautology of extraction across pulse and filtered Gaussian noise disturbance types.} Intent is modeled as a minimum jerk, 5th order polynomial. Forces experienced are combined with intent via Burdet et al.'s \cite{burdet2006stability} model to produce the simulated arm trajectory. Extraction to recover intention from arm and force trajectory follows. Parameter errors are introduced into the extraction and varied according to Table 1 to estimate sensitivity. Examining the distribution of deviation in trajectories from this analysis reveals that sensitivity is not uniformly distributed throughout a reach.
}
\label{fig:synthetic}
\end{figure}

\begin{figure}
\caption{
{\bf Model sensitivity testing (from the variance-based method of Saltelli et al.\cite{saltelli2010variance}) reveals that the model is mostly sensitive to stiffness.} Parameter values are varied in a Sobol-distributed fashion according to Table 1. First order sensitivity, in purple, is the error that would be removed if the parameter was fixed at its nominal value. Total sensitivity, in teal, shows error that would remain if all other parameters were fixed at their nominal values.
}
\label{fig:sensitivity}
\end{figure}

\begin{figure}
\caption{
{\bf Subjects' disturbed hand trajectories and extraction of intended trajectories from them reveal that both variance and invariance of the motor plan can occur even in response to very large disturbances.} The hand path, in black, deviates from the blue baseline, as a force pulse, gray arrows, is applied. Three of the eight subjects' intent trajectories (orange) did not significantly deviate from undisturbed movement (panel A) as observed by the intent region not departing the undisturbed movement (blue) region. Shaded regions represent mean $\pm$ 95\% confidence interval on the mean. The first time at which the hand (black) region departed from undisturbed movement is marked with a green dot for each subject. Five of the eight subjects' intent trajectories did significantly deviate from undisturbed movement (panel B). The time of intent departure from the undisturbed trajectories is marked with a red dot for each subject.
}
\label{fig:anecdotes}
\end{figure}
\pagebreak

>>>>

\usepackage[section]{placeins}
\usepackage{graphics} % for pdf, bitmapped graphics files

@@

\subsection*{Intent Extraction: Deducing the Desired Trajectory}
We first demonstrate the \textit{intent extraction} approach in human motor control, but we later show that the process is applicable to any controlled process with an invertible highest order plant term. The process begins by presuming a model of the controller. Here, we choose the well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} where the feedforward aspect of the controller perfectly predicts the plant and linearizes the system through cancellation. Additionally, linear feedback rejects position and velocity error. This model was chosen to help illuminate the approach, but later we present how to generalize this approach to more complex models without loss of generality. The equation governing the passive planar dynamics (plant) of musculoskeletal structure is of the form,
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripetal}}}_{\text{Plant}}+E=0
\end{equation}
where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocity, $\ddot{q}$ is joint angular acceleration, $G$ contains both Coriolis and centripetal effects, and $E$ is any externally-applied torque. The motion behavior changes with the addition of feedforward and/or feedback controllers,
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripetal}}}_{\text{Plant}}+E=\underbrace{\overbrace{\hat{M}(q_d)\ddot{q}_d}^{\text{Inertia}}+\overbrace{\hat{G}(q_d,\dot{q}_d)}^{\text{Coriolis, Centripetal}}+\hat{E}}_{\text{Feedforward Controller}}+\underbrace{K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})}_{\text{Impedance, Feedback Controller}}
\end{equation}
where terms with hats over them ($\hat{M}$, $\hat{G}$, or $\hat{E}$) indicate that they represent the nervous system's best estimate of the forces and dynamics it will encounter, which is also known as an internal model \cite{shadmehr1994adaptive}. This portion of the system serves as an inverse-dynamics feedforward controller that cancels out the dynamics of the arm in the torque balance. If the nervous system has sufficient experience and is expecting $E$, it is included as part of the internal model, $\hat{E}$; otherwise $\hat{E}$ is set to zero. $K_p$ and $K_d$ are the lumped impedance and feedback terms that employ a moving state equilibrium to accomplish the desired trajectory, $q_d$. This $q_d$ has a dual meaning in that it signifies both the unknown desired trajectory that we seek to discover and also the moving equilibrium trajectory of the arm.

@@

To explore the conditions under which this procedure for recovering the intent trajectory can be generalized, we take a series of steps. First, we distinguish between joint space and generalized space by introducing a generalized state variable, $x$. Second, we separate the plant (arm) into its process (dynamics), $P$, and actuator (muscles), $A$, components. Third, we model these components in a very general sense as operations,
\begin{equation} \label{eq:generalPlant}
\underbrace{\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+\overbrace{\sum_{m=1}^M A_m(x-x_e)^{(m)}}^\text{Actuator}}_\text{Plant}+E=0
\end{equation}
where $x_e$ is the equilibrium trajectory of the actuator. Fourth, we describe a  control law for the actuator using the same expansions,
\begin{equation}
\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m(x_d-x_e)^{(m)}}^\text{Actuator}=0
\end{equation}
where an internal model, $\hat{P}$ and $\hat{E}$, predicts system dynamics and external disturbances in order to determine the actuator equilibrium $x_e$ such that $x$ will track a desired path $x_d$. The feedforward component must use the actuator and hence it shares the actuator's equilibrium with the plant. This allows us to finally relate the physical system and its control law by solving for $x_e$ and substituting into \eqref{eq:generalPlant},
\begin{equation}
\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+E=\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m (x_d-x)^{(m)}}^\text{Feedback (Actuator)}
\end{equation}
wherein $x_e$ vanishes recovering our familiar model. While a proper choice of $x_e$ is perhaps essential for control, our approach does not require a model relating this equilibrium and intended trajectory. Note also that the actuator's equilibrium $x_e$ is not the process's equilibrium unless all derivatives of $x_d$ are zero.

The fact that the actuator's equilibrium $x_e$ is not the process's equilibrium is important as the actuator state ($\lambda$) has been claimed to be the intended movement\cite{feldman1995origin}. This upholds Gomi and Kawato's finding that the arm muscles' equilibrium does not represent reaching intent\cite{gomi1997human}. If the highest order coefficient of the dynamic model of the process $\hat{P}_N$ can be inverted, the impedances can be modeled, and the system is stable (e.g., the stiffness and damping are positive), it is possible to solve for $x_d^{(N)}$ and integrate for $x_d$, revealing the intended trajectory.

While we treat both the process and actuator as dynamic models in our derivation, it is important to note that both the derivation and the technique are agnostic to the process model, simply because any hypothesis can be modeled and used by our approach to determine where the model would have gone had it not been disturbed.

Note that instead of a dynamic model, any model may be possible, such as lookup table. For instance, a switch is well-modeled as a lookup table or threshold without any consideration of its underlying mechanism. In this case, our technique could be used to determine a person's intention to flip the switch and how it changes in the face of disturbance. As long as some bidirectional relationship exists between state and outcome (even if determined empirically), intended outcome can be determined. This allows determination of intent in many situations of interest, even where the process is otherwise irreducibly complex.

@@

\subsection*{Indices of variance-based sensitivity}
To ensure a properly spaced and efficient evaluation of sensitivity, we employed Sobol-distributed matrices, generated using MATLAB's \textit{sobolset()} function and converted to parameter distributions (see Table~\ref{tab:parameters}) using inverse cumulative probability density. Nominal parameter values were mostly taken from Dempster \cite{dempster1955space} and Burdet et al. \cite{burdet2006stability} with typical values for subject-specific parameters (height and weight) and direct measurements for device-specific parameters (force sensor noise and drift). Standard deviations reflect expected variation in repeated measurement (ie. height and weight) or measured variance (ie. trial-to-trial variation in sensor noise or uncertainty in mass ratios reported by Dempster \cite{dempster1955space}). Distributions were clamped to the range of $\pm3$ standard deviations in order to limit parameters to a realistic range. Combinations and calculations were made as prescribed by Saltelli et al. \cite{saltelli2010variance} to arrive at direct and total sensitivity indices.

@@

Nevertheless, there are some limitations to these methods. Accurate interface force measurement (within a few tenths of a newton) is needed. Such instrumentation is available, but must be used with great care in order to preserve the accuracy of the estimate. Our sensitivity analysis on synthetic data revealed that slight error can lead to a trajectory that accumulates a drift due to bias caused by the force term. More importantly, while modern force sensors can be highly accurate with very high signal to noise ratios, they tend to drift over time, leading to error that can grow if the device is not periodically tared. Our empirical approach used several methods to mitigate these effects, and hence the errors became small. It remains to be seen if there are methods to further reduce the mis-estimation due to force error or eliminate the need for force estimation altogether. It will be also interesting to determine whether relaxing constraints and operating in 3D would produce the same levels of accuracy found here.

A key assumption in this study is that the joint impedance (stiffness) is linearly related to torque. While other approaches tend to construct paradigms that indicate or presume that the intention is static in order to derive impedance -- the so called family of system identification applications, this study makes a constrained assumption of stiffness in order to determine intent. In a sense, this approach widens the scope to include movement. Sensitivity analyses showed that approximation of stiffness is the most critical factor to accuracy in estimating intent. While serious trial-to-trial stiffness mis-estimation is unlikely, co-contraction in response to disturbance might plausibly double stiffness mid-reach. Even as the model is insensitive to normal variations, a human can halve or double their stiffness \cite{franklin2007endpoint} significantly reducing determination accuracy. This stiffening may even explain the reported deviation of intent from the baseline trajectories. Of particular concern are the large hooks at the end of the motions. These limitations may be overcome by detecting muscle co-contraction or using additional modeling assumptions to allow real-time estimation of stiffness.

@@

% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgements}
Research supported by US NIH NINDS1 R01 NS053606.  We thank Felix Huang, Jagan Jimmy, Christine Massie, Tejas Madhavan, and Mohi Ahmed for their review of preliminary manuscript. We also thank the community  of the Robotics Lab at the Rehabilitation Institute of Chicago for critical commentary during the genesis of this paper and the work leading up to it.

%\section*{References}
\bibliography{paper1.bib}
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% OR
%
% Compile your BiBTeX database using our plos2009.bst
% style file and paste the contents of your .bbl file
% here.
%

\pagebreak

\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}[!b]
\includegraphics{fig1.eps}
\caption{
{\bf Simulated data illustrating tautology of extraction across pulse and filtered Gaussian noise disturbance types.} Intent is modeled as a minimum jerk, 5th order polynomial. Forces experienced are combined with intent via Burdet et al.'s \cite{burdet2006stability} model to produce the simulated arm trajectory. Extraction to recover intention from arm and force trajectory follows. Parameter errors are introduced into the extraction and varied according to Table 1 to estimate sensitivity. Examining the distribution of deviation in trajectories from this analysis reveals that sensitivity is not uniformly distributed throughout a reach.
}
\label{fig:synthetic}
\end{figure}

\begin{figure}
\includegraphics{fig2.eps}
\caption{
{\bf Model sensitivity testing (from the variance-based method of Saltelli et al.\cite{saltelli2010variance}) reveals that the model is mostly sensitive to stiffness.} Parameter values are varied in a Sobol-distributed fashion according to Table 1. First order sensitivity, in purple, is the error that would be removed if the parameter was fixed at its nominal value. Total sensitivity, in teal, shows error that would remain if all other parameters were fixed at their nominal values.
}
\label{fig:sensitivity}
\end{figure}

\begin{figure}
\includegraphics{fig3.eps}
\caption{
{\bf Subjects' disturbed hand trajectories and extraction of intended trajectories from them reveal that both variance and invariance of the motor plan can occur even in response to very large disturbances.} The hand path, in black, deviates from the blue baseline, as a force pulse, gray arrows, is applied. Three of the eight subjects' intent trajectories (orange) did not significantly deviate from undisturbed movement (panel A) as observed by the intent region not departing the undisturbed movement (blue) region. Shaded regions represent mean $\pm$ 95\% confidence interval on the mean. The first time at which the hand (black) region departed from undisturbed movement is marked with a green dot for each subject. Five of the eight subjects' intent trajectories did significantly deviate from undisturbed movement (panel B). The time of intent departure from the undisturbed trajectories is marked with a red dot for each subject.
}
\label{fig:anecdotes}
\end{figure}
\pagebreak
